Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2015-06-18 cordillera-cycle.tex
Update Martin's affiliation.
Add corrections from Irina (part 1).

% Authors
\Author[1,2,3]{J.}{Seguinot}
\Author[3]{I.}{Rogozhina}
\Author[2]{A.~P.}{Stroeven}
\Author[4]{M.}{Margold}
\Author[2]{J.}{Kleman}
\runningauthor{J.~Seguinot et~al.}
\correspondence{J.~Seguinot (seguinot@vaw.baug.ethz.ch)}

@@

% Affiliations
\affil[1]{Laboratory of Hydraulics, Hydrology and Glaciology, ETH Zürich,
          Zürich, Switzerland}
\affil[2]{Department of Physical Geography and the Bolin Centre for Climate
          Research, Stockholm University, Stockholm, Sweden}
\affil[3]{Helmholtz Centre Potsdam, GFZ German Research Centre for Geosciences,
          Potsdam, Germany}
\affil[4]{Department of Geography, Durham University, UK}

@@

More than a century of exploration and geological investigation of the
Cordilleran mountains have led to many observations in support of the former
ice sheet
    \citep{Jackson.Clague.1991}.
Despite the lack of documented end moraines offshore, in the zone of confluence
with the Laurentide ice sheet, and in areas swept by the Missoula floods
    \citep{Carrara.etal.1996},
moraines that demarcate the northern and south-western margins provide key
constraints that allow reasonable reconstructions of maximum ice sheet extents
    (\citealp{Prest.etal.1968}; \citealp[Fig. 1.12]{Clague.1989};
     \citealp{Duk-Rodkin.1999};
     \citealp{Booth.etal.2003}; \citealp{Dyke.2004}).
The LGM Cordilleran ice sheet extent was short-lived, as indicated by
field evidence from radiocarbon dating
    \citep{Clague.etal.1980, Clague.1985, Clague.1986, Porter.Swanson.1998,
           Menounos.etal.2008},
cosmogenic exposure dating
    \citep{Stroeven.etal.2010, Stroeven.etal.2014, Margold.etal.2014},
bedrock deformation in response to former ice loads
    \citep{Clague.James.2002, Clague.etal.2005},
and offshore sedimentary records
    \citep{Cosma.etal.2008, Davies.etal.2011}.
However, former ice thicknesses and, therefore, the ice sheet's contribution to
the LGM sea level lowstand
    \citep{Carlson.Clark.2012, Clark.Mix.2002}
remain uncertain.

Our understanding of the Cordilleran glaciation history prior to the LGM is
even more fragmentary
    \citep{Barendregt.Irving.1998, Kleman.etal.2010, Rutter.etal.2012},
although it is clear that maximum glaciation of the Cordilleran ice sheet
predates the last glacial cycle
    \citep{Hidy.etal.2013}.
In parts of the Yukon Territory and Alaska, the distribution of tills
    \citep{Turner.etal.2013}
and dated glacial erratics
    \citep{Ward.etal.2007, Ward.etal.2008, Briner.Kaufman.2008,
           Stroeven.etal.2010, Stroeven.etal.2014}
indicate an extensive Marine Oxygen Isotope Stage (MIS)~4 glaciation, of which
evidence have also been found in the Puget Lowland \citep{Troost.2014}.
Landforms in the interior regions include flow sets that are likely
older than the LGM
    \citep[Fig.~2]{Kleman.etal.2010},
but their absolute age remains uncertain.

@@

\begin{itemize}
  \item How much ice was locked in the Cordilleran ice sheet during the
    last glacial maximum?
  \item What was the scale of glaciation prior to the last glacial maximum?
  \item Which were the primary dispersal centres? Do they reflect stable or
    ephemeral configurations?
  \item How rapid was the last deglaciation? Did it include late glacial
    standstills or readvances?
\end{itemize}

Although numerical ice sheet modelling has been established as a useful tool to
improve our understanding of the Cordilleran ice sheet
    (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991},
     \citealp{Marshall.etal.2000}),
the ubiquitously mountainous
topography of the region has presented two major challenges to its application.
First, only recent development of numerical ice sheet models and underlying
scientific computing tools \citep{Bueler.Brown.2009, Balay.etal.2015} has
allowed for high-resolution numerical modelling of glaciers and ice sheets on
mountainous terrain
over millennial time scales \citep[e.g.,][]{Golledge.etal.2012}. Second, the
complex
topography of the North American Cordillera also induces strong gradients in
temperature
seasonality and distinct patterns of precipition, thus requiring the use of
high-resolution climate forcing fields as an input to the ice sheet model
\citep{Seguinot.etal.2014}.

Because climate conditions over the last glacial cycle are subject to
considerable uncertainty, our palaeo-climate forcing is a simplistic approach
including temperature and precipitation fields derived from a
present-day atmospheric reanalysis \citep{Mesinger.etal.2006,
Seguinot.etal.2014} supplemented by lapse-rate corrections
and temperature offset time series. The latter are obtained by scaling six
different palaeo-temperature reconstructions from proxy records around the
globe, including two oxygen isotope records from Greenland ice cores
\citep{Dansgaard.etal.1993, Andersen.etal.2004}, two oxygen isotope
records from Antarctic ice cores \citep{Petit.etal.1999,Jouzel.etal.2007},
and two alkenone unsaturation index records from Northwest Pacific ocean
sediment cores \citep{Herbert.etal.2001}. We then proceed to compare the model
output to geological evidence and discuss the timing and extent of glaciation
and the patterns of deglaciation.

@@

Basal topography is derived from the ETOPO1 combined topography and bathymetry
dataset with a~resolution of 1\,arc-min \citep{Amante.Eakins.2009}. Sea level
is lowered as a function of time based on the Spectral Mapping Project
\citep[SPECMAP,][]{Imbrie.etal.1989} time scale. Geothermal heat flux
is applied as a constant value of 70\,\unit{mW\,m^{-2}} at 3\,km depth
(Sect.~\ref{sec:icedyn}). Surface mass balance is computed using a positive
degree-day (PDD) model (Sect.~\ref{sec:surface}). Climate forcing is
provided by a monthly climatology averaged from 1979 to 2000 from the North
American Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006},
perturbated by time-dependent and
lapse-rate temperature corrections (Sect.~\ref{sec:atm}).

@@

Ice sheet dynamics are typically modelled using a combination of internal
deformation and basal sliding. PISM is a~shallow ice sheet model, which implies
that the balance of stresses is approximated based on their predominant
components. The Shallow Shelf Approximation (SSA) is used as a ``sliding law''
for the Shallow Ice Approximation (SIA) by adding velocity solutions of the
two approximation
\citep[Eqns.~7--9 and 15]{Bueler.Brown.2009, Winkelmann.etal.2011}.

Ice rheology depends on temperature and water content through an enthalpy
formulation \citep{Aschwanden.etal.2012}. Surface air temperature derived from
the climate forcing (Sect.~\ref{sec:atm}) provides the upper boundary
condition to the ice enthalpy model. Temperature is computed subglacially to
a~depth of 3\,km,
where it is conditioned by a lower boundary geothermal heat flux of
70\,\unit{mW\,m^{-2}}. Although this uniform value does not account for the
high spatial geothermal variability in the region
\citep{Blackwell.Richards.2004}, it is, on average, representative of available
heat flow measurements. In the low-resolution simulations, the vertical grid
consists of up to 51~enthalpy layers in the ice sheet and 31~temperature layers
in the bedrock. In the high-resolution simulations, up to 101~ice layers, and
61~bedrock layers are used.

A~pseudo-plastic sliding law,
\begin{equation}
    \label{eqn:pseudoplastic}
    \vec{\tau}_b = -\tau_c \frac{\vec{v}_b}{{v_{th}}^q\,|\vec{v}_b|^{1-q}} \,,
\end{equation}
relates the bed-parallel shear stresses, $\vec{\tau}_b$, to the sliding
velocity, $\vec{v}_b$ (Table~\ref{tab:params}).
The yield stress, $\tau_c$,
is modelled using the Mohr--Coulomb criterion,
\begin{equation}
   \tau_c = c_0 + N\,\tan{\phi} \,,
\end{equation}
where cohesion, $c_0$, is assumed to be zero. The friction angle, $\phi$,
varies from 15 to 45\degree as a piecewise-linear function of modern bed
elevation, with the lowest value occuring below modern sea level (0\,m above
sea level, m~a.s.l.) and the highest value occuring above the generalised
elevation of the highest shorelines
\citep[200\,m~a.s.l.,][Fig.~5]{Clague.1981}, thus accounting for
a~weakening of till associated with the presence of marine sediments
(cf. \citealp{Martin.etal.2011}; \citealp[supplement]{Aschwanden.etal.2013};
\citealp{PISM-authors.2015}). Effective pressure, $N$,
is related to the ice overburden stress, $\rho gh$, and the modelled amount of
subglacial water, using a formula derived from laboratory experiments with till
extracted from an Antarctic ice stream \citep{Tulaczyk.etal.2000,
Bueler.Pelt.2014},
\begin{equation}
    N = \delta \rho gh \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,,
\end{equation}
where $\delta$ is chosen as 0.02, $e_0$ is a measured reference void ratio and
$C_c$ a measured compressibility coefficient (Table~\ref{tab:params}). The
amount of water at the base, $W$, varies from zero to $W_{max}=2$\,m, a
threshold above which any additional melt water is assumed to drain off
instantaneously. Finally, the bedrock topography responds to ice load
following a bedrock deformation model that includes local isostasy,
elastic lithosphere flexure and viscous astenosphere deformation in an infinite
half-space
\citep[Table~\ref{tab:params};][]{Lingle.Clark.1985,Bueler.etal.2007}.
A relatively low viscosity value of $\nu_m = 1\times10^{19}$\,\unit{Pa\,s} is
used for the astenosphere in accordance to the results of regional glacial
isostatic adjustment modelling at the northern Cascadia subduction zone
\citep{James.etal.2009}.

Ice-shelve calving is computed by a double criterion. First, a
physically-realistic calving flux is computed based on eigenvalues of the
horizontal strain rate tensor \citep{Winkelmann.etal.2011,
Levermann.etal.2012}. This allows outlet glaciers to advance in confined
embayments, but prevents extensive ice-shelves to build-up in the open ocean.
Second, floating ice thinner than 50\,m is systematically calved off. A subgrid
scheme \citep{Albrecht.etal.2011} allows for a continuous migration of the
calving front. Such treatment of calving was applied to the Antarctic ice
sheet and shown to produce realistic calving front positons for many of the
ice-shelves \citep{Martin.etal.2011}.

@@

Ice surface accumulation and ablation are computed from monthly mean
near-surface air temperature, $T_m$, monthly standard deviation of near-surface
air temperature, $\sigma$, and monthly precipitation, $P_m$, using
a~temperature-index model \citep[e.g.,][]{Hock.2003}.
Accumulation is equal to
precipitation when air temperatures are below 0\,\unit{{\degree}C}, and
decreases to zero linearly with temperatures between 0 and
2\,\unit{{\degree}C}. Ablation is computed from the PDD, defined as the
integral of temperatures above 0\,\unit{{\degree}C} in one year.

The PDD computation accounts for stochastic temperature variations by assuming
a normal temperature distribution of standard deviation $\sigma$ aroung the
expected value $T_m$. It is expressed by an error-function formulation
\citep{Calov.Greve.2005},
\begin{equation}
    \label{eqn:calovgreve}
    \mathrm{PDD} = \int_{t_1}^{t_2} \mathrm{d}t
        \left[\frac{\sigma}{\sqrt{2\pi}}
                \exp\left({-\frac{T_{m}^2}{2\sigma^2}}\right)
              + \frac{T_{m}}{2} \, \mathrm{erfc}
                \left(-\frac{T_{m}}{\sqrt{2}\sigma}\right)\right] \,,
\end{equation}
which is numerically approximated using week-long sub-intervals. In order to
account for the effects of spatial and seasonal variations of temperature
variability \citep{Seguinot.2013}, $\sigma$ is computed from daily temperature
values from the NARR, including variability associated with the seasonal cycle
itself \citep[Fig.~\ref{fig:atm}; i.e. not as in][]{Seguinot.Rogozhina.2014}.
Degree-day factors for snow and ice melt are derived from
mass-balance measurements on contemporary glaciers from the Coast Mountains and
Rocky Mountains in British Columbia
\citep[Table~\ref{tab:params};][]{Shea.etal.2009}.

@@

Climate forcing of the model consists of a present-day monthly climatology,
$\{T_{m0}, P_{m0}\}$, where temperatures are modified by offset time series,
${\Delta}T_{TS}$, and lapse-rate corrections, ${\Delta}T_{LR}$:
\begin{align}
    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{TS}(t)
                    + {\Delta}T_{LR}(t, x, y) \,, \\
    P_m(t, x, y) &= P_{m0}(x, y) \,.
\end{align}

@@

Despite large differences in the input climate forcing
(Fig.~\ref{fig:timeseries}, upper panel), model output presents consistent
features that can be observed across the range of forcing data used. In all
simulations, modelled ice volumes remain relatively low during most of the
glacial cycle, except during two major glacial events which occur between 61.9
and 56.5\,ka during MIS~4, and between 23.2 and 16.8\,ka during MIS~2
(Fig.~\ref{fig:timeseries}, lower panel). An ice volume minimum is
consistently reached between 53.0 and 41.3\,ka during MIS~3. However, the
magnitude and precise timing of these three events depend significantly on the
choice of proxy record used for time-dependent climate forcing
(Table~\ref{tab:extrema}).

Simulations forced by the Greenland ice core palaeo-temperature
records (GRIP, NGRIP) produce the highest variability in modelled ice volume
throughout the last glacial cycle. In contrast, simulations driven by oceanic
(ODP~1012, ODP~1020) and Antarctic (EPICA, Vostok) palaeo-temperature records
generally result in smaller
modelled ice volumes during MIS~4 and larger ice volumes during MIS~3.
Hence, they produce lower ice volume variability
throughout the simulation length. The NGRIP climate forcing is the only one
that results in a larger ice volume during MIS~4 than during MIS~2.

While simulations driven by the GRIP and the two Antarctic palaeo-temperature
records attain a last ice volume maximum between 19.1 and 16.8\,ka, those
informed by the NGRIP and the two oceanic palaeo-temperature records attain their
maximum ice volumes thousands of years earlier. Moreover, the ODP~1012 run
yields a rapid deglaciation of the modelled area prior to 17\,ka. The ODP~1020
simulation predicts an early maximum in ice volume at 23.2\,ka, followed by
slower deglaciation than modelled using the other palaeo-temperature records.
Finally, model runs forced by Antarctic palaeo-temperature records result in a
rapid and uninterrupted deglaciation after the last glacial maximum, whereas the
simulation driven by the GRIP palaeo-temperature record also results in a rapid
deglaciation but in three steps, separated by two periods of ice volume increase
(Fig.~\ref{fig:timeseries}).

@@

Despite such differences in the timing of attained volume extrema
(Table~\ref{tab:extrema}), all model runs show relatively consistent
patterns of glaciation. During MIS~4, all simulations produce an extensive ice
sheet, covering an area of at least half of that attained during MIS~2
(Table~\ref{tab:extrema}; Fig.~\ref{fig:snapshots}, upper panels).
Corresponding maximum ice volumes also differ significantly between model runs,
and vary between 3.84 and~8.84\,m sea level equivalents (m.~s.l.e.;
Table~\ref{tab:extrema}).

During MIS~3 ice volume minima, a central ice cap persists over the Skeena
Mountains (Fig.~\ref{fig:snapshots}, middle panels). Although this ice cap is
present in all simulations, its dimensions depend sensitively on the choice of
the applied palaeo-temperature record. Modelled ice volume minima spread over a
wide range between 1.69 and 2.88\,m~s.l.e. (Table~\ref{tab:extrema}).

Modelled ice sheet geometries during the LGM (MIS~2; Fig.~\ref{fig:snapshots},
lower panels) invariably include a ca. 1500\,km-long central divide located
about 3500\,m~a.s.l. along the spine of the Rocky Mountains. Although the
similatity of modelled ice extents is a direct result from the choice of scaling
factors applied to different palaeo-temperature proxy records
(Table~\ref{tab:records}), it is interesting to note that modelled maximum ice
volumes also fall within a tight range of 8.40 to 8.91\,m~s.l.e.
(Table~\ref{tab:extrema}).

>>>>

% Authors
\Author[1,2,3]{J.}{Seguinot}
\Author[3]{I.}{Rogozhina}
\Author[2]{A.~P.}{Stroeven}
\Author[2]{M.}{Margold}
\Author[2]{J.}{Kleman}
\runningauthor{J.~Seguinot et~al.}
\correspondence{J.~Seguinot (seguinot@vaw.baug.ethz.ch)}

@@

% Affiliations
\affil[1]{Laboratory of Hydraulics, Hydrology and Glaciology, ETH Zürich,
          Zürich, Switzerland}
\affil[2]{Department of Physical Geography and the Bolin Centre for Climate
          Research, Stockholm University, Stockholm, Sweden}
\affil[3]{Helmholtz Centre Potsdam, GFZ German Research Centre for Geosciences,
          Potsdam, Germany}

@@

More than a century of exploration and geological investigation of the
Cordilleran mountains have led to many observations in support of the former
ice sheet
    \citep{Jackson.Clague.1991}.
Despite the lack of documented end moraines offshore, in the zone of confluence
with the Laurentide ice sheet, and in areas swept by the Missoula floods
    \citep{Carrara.etal.1996},
moraines that demarcate the northern and south-western margins provide key
constraints that allow reasonable reconstructions of maximum ice sheet extents
    (\citealp{Prest.etal.1968}; \citealp[Fig. 1.12]{Clague.1989};
     \citealp{Duk-Rodkin.1999};
     \citealp{Booth.etal.2003}; \citealp{Dyke.2004}).
As indicated by field evidence from radiocarbon dating
    \citep{Clague.etal.1980, Clague.1985, Clague.1986, Porter.Swanson.1998,
           Menounos.etal.2008},
cosmogenic exposure dating
    \citep{Stroeven.etal.2010, Stroeven.etal.2014, Margold.etal.2014},
bedrock deformation in response to former ice loads
    \citep{Clague.James.2002, Clague.etal.2005},
and offshore sedimentary records
    \citep{Cosma.etal.2008, Davies.etal.2011},
the LGM Cordilleran ice sheet extent was short-lived.
However, former ice thicknesses and, therefore, the ice sheet's contribution to
the LGM sea level lowstand
    \citep{Carlson.Clark.2012, Clark.Mix.2002}
remain uncertain.

Our understanding of the Cordilleran glaciation history prior to the LGM is
even more fragmentary
    \citep{Barendregt.Irving.1998, Kleman.etal.2010, Rutter.etal.2012},
although it is clear that maximum glaciation of the Cordilleran ice sheet
predates the last glacial cycle
    \citep{Hidy.etal.2013}.
In parts of the Yukon Territory and Alaska, the distribution of tills
    \citep{Turner.etal.2013}
and dated glacial erratics
    \citep{Ward.etal.2007, Ward.etal.2008, Briner.Kaufman.2008,
           Stroeven.etal.2010, Stroeven.etal.2014}
indicate an extensive Marine Oxygen Isotope Stage (MIS)~4 glaciation, the
evidence of which have also been found in the Puget Lowland \citep{Troost.2014}.
Landforms in the interior regions include flow sets that are likely
older than the LGM
    \citep[Fig.~2]{Kleman.etal.2010},
but their absolute age remains uncertain.

@@

\begin{itemize}
  \item How much ice was locked in the Cordilleran ice sheet during the LGM?
  \item What was the scale of glaciation prior to the LGM?
  \item Which were the primary dispersal centres? Do they reflect stable or
    ephemeral configurations?
  \item How rapid was the last deglaciation? Did it include late glacial
    standstills or readvances?
\end{itemize}

Although numerical ice sheet modelling has been established as a useful tool to
improve our understanding of the Cordilleran ice sheet
    (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991},
     \citealp{Marshall.etal.2000}),
the ubiquitously mountainous
topography of the region has presented two major challenges to its application.
First, only recent development of numerical ice sheet models and underlying
scientific computing tools \citep{Bueler.Brown.2009, Balay.etal.2015} has
allowed for high-resolution numerical modelling of glaciers and ice sheets on
mountainous terrain
over millennial time scales \citep[e.g.,][]{Golledge.etal.2012}. Second, the
complex topography of the North American Cordillera also induces strong
geographic variations in temperature and precipitation, thus requiring the use
of high-resolution climate forcing fields as an input to an ice sheet model
\citep{Seguinot.etal.2014}. However, evolving climate conditions over the last
glacial cycle are subject to considerable uncertainty and still lie beyond the
computational reach of atmosphere circulation models.

Our palaeo-climate forcing therefore includes spatial temperature and
precipitation grids derived from a present-day atmospheric reanalysis
\citep{Mesinger.etal.2006} that includes the steep precipitation gradients
previously identified as necessary to model the LGM extent of the Cordilleran
ice sheet in agreement with its geological imprint
\citep{Seguinot.etal.2014}. To mimic climate evolution through the last glacial
cycle, these grids are simply supplemented by lapse-rate corrections
and temperature offset time series. The latter are obtained by scaling six
different palaeo-temperature reconstructions from proxy records around the
globe, including two oxygen isotope records from Greenland ice cores
\citep{Dansgaard.etal.1993, Andersen.etal.2004}, two oxygen isotope
records from Antarctic ice cores \citep{Petit.etal.1999,Jouzel.etal.2007},
and two alkenone unsaturation index records from Northwest Pacific ocean
sediment cores \citep{Herbert.etal.2001}. We then proceed to compare the model
output to geological evidence and discuss the timing and extent of glaciation
and the patterns of deglaciation, based on which we use the applicability of
different records to modelling the history of the Cordilleran ice sheet.

@@

Basal topography is derived from the ETOPO1 combined topography and bathymetry
dataset with a~resolution of 1\,arc-min \citep{Amante.Eakins.2009}. Sea level
is lowered as a function of time based on the Spectral Mapping Project
\citep[SPECMAP,][]{Imbrie.etal.1989} time scale. Geothermal heat flux
is applied as a constant value of 70\,\unit{mW\,m^{-2}} at 3\,km depth
(Sect.~\ref{sec:icedyn}). Surface mass balance is computed using a positive
degree-day (PDD) model (Sect.~\ref{sec:surface}). Climate forcing is
provided by a monthly climatology averaged from 1979 to 2000 from the North
American Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006},
perturbated by time-dependent offsets and
lapse-rate temperature corrections (Sect.~\ref{sec:atm}).

@@

Ice sheet dynamics are typically modelled using a combination of internal
deformation and basal sliding. PISM is a~shallow ice sheet model, which implies
that the balance of stresses is approximated based on their predominant
components. The Shallow Shelf Approximation (SSA) is used as a ``sliding law''
for the Shallow Ice Approximation (SIA) by adding velocity solutions of the
two approximations
\citep[Eqns.~7--9 and 15]{Bueler.Brown.2009, Winkelmann.etal.2011}.

Ice rheology depends on temperature and water content through an enthalpy
formulation \citep{Aschwanden.etal.2012}. Surface air temperature derived from
the climate forcing (Sect.~\ref{sec:atm}) provides the upper boundary
condition to the ice enthalpy model. Temperature is computed subglacially to
a~depth of 3\,km,
where it is conditioned by a lower boundary geothermal heat flux of
70\,\unit{mW\,m^{-2}}. Although this uniform value does not account for the
high spatial geothermal variability in the region
\citep{Blackwell.Richards.2004}, it is, on average, representative of available
heat flow measurements. In the low-resolution simulations, the vertical grid
consists of up to 51~enthalpy layers in the ice sheet and 31~temperature layers
in the bedrock. In the high-resolution simulations, up to 101~ice layers and
61~bedrock layers are used.

A~pseudo-plastic sliding law,
\begin{equation}
    \label{eqn:pseudoplastic}
    \vec{\tau}_b = -\tau_c \frac{\vec{v}_b}{{v_{th}}^q\,|\vec{v}_b|^{1-q}} \,,
\end{equation}
relates the bed-parallel shear stresses, $\vec{\tau}_b$, to the sliding
velocity, $\vec{v}_b$ (Table~\ref{tab:params}).
The yield stress, $\tau_c$,
is modelled using the Mohr--Coulomb criterion,
\begin{equation}
   \tau_c = c_0 + N\,\tan{\phi} \,,
\end{equation}
where cohesion, $c_0$, is assumed to be zero. The friction angle, $\phi$,
varies from 15 to 45\degree as a piecewise-linear function of modern bed
elevation, with the lowest value occuring below modern sea level (0\,m above
sea level, m~a.s.l.) and the highest value occuring above the generalised
elevation of the highest shorelines
\citep[200\,m~a.s.l.,][Fig.~5]{Clague.1981}, thus accounting for
a~weakening of till associated with the presence of marine sediments
(cf. \citealp{Martin.etal.2011}; \citealp[supplement]{Aschwanden.etal.2013};
\citealp{PISM-authors.2015}). Effective pressure, $N$,
is related to the ice overburden stress, $\rho gh$, and the modelled amount of
subglacial water, using a formula derived from laboratory experiments with till
extracted from the base of Ice Stream B in West Antarctica
\citep{Tulaczyk.etal.2000, Bueler.Pelt.2014},
\begin{equation}
    N = \delta \rho gh \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,,
\end{equation}
where $\delta$ is chosen as 0.02, $e_0$ is a measured reference void ratio and
$C_c$ is a measured compressibility coefficient (Table~\ref{tab:params}). The
amount of water at the base, $W$, varies from zero to $W_{max}=2$\,m, a
threshold above which any additional melt water is assumed to drain off
instantaneously. Finally, the bedrock topography responds to ice load
following a bedrock deformation model that includes local isostasy,
elastic lithosphere flexure and viscous astenosphere deformation in an infinite
half-space
\citep[Table~\ref{tab:params};][]{Lingle.Clark.1985,Bueler.etal.2007}.
A relatively low viscosity value of $\nu_m = 1\times10^{19}$\,\unit{Pa\,s} is
used for the astenosphere in accordance to the results from regional glacial
isostatic adjustment modelling at the northern Cascadia subduction zone
\citep{James.etal.2009}.

Ice-shelf calving is computed by a double criterion. First, a
physically-realistic calving flux is computed based on eigenvalues of the
horizontal strain rate tensor \citep{Winkelmann.etal.2011,
Levermann.etal.2012}. This allows outlet glaciers to advance in confined
embayments, but prevents extensive ice-shelves from building up in the open ocean.
Second, floating ice thinner than 50\,m is systematically calved off. A subgrid
scheme \citep{Albrecht.etal.2011} allows for a continuous migration of the
calving front. Such treatment of calving has been applied to the Antarctic ice
sheet and has shown to produce a realistic calving front positon for many of
the present-day ice shelves \citep{Martin.etal.2011}.

@@

Ice surface accumulation and ablation are computed from monthly mean
near-surface air temperature, $T_m$, monthly standard deviation of near-surface
air temperature, $\sigma$, and monthly precipitation, $P_m$, using
a~temperature-index model \citep[e.g.,][]{Hock.2003}.
Accumulation is equal to
precipitation when air temperatures are below 0\,\unit{{\degree}C}, and
decreases to zero linearly with temperatures between 0 and
2\,\unit{{\degree}C}. Ablation is computed from PDD, defined as an
integral of temperatures above 0\,\unit{{\degree}C} in one year.

The PDD computation accounts for stochastic temperature variations by assuming
a normal temperature distribution of standard deviation $\sigma$ aroung the
expected value $T_m$. It is expressed by an error-function formulation
\citep{Calov.Greve.2005},
\begin{equation}
    \label{eqn:calovgreve}
    \mathrm{PDD} = \int_{t_1}^{t_2} \mathrm{d}t
        \left[\frac{\sigma}{\sqrt{2\pi}}
                \exp\left({-\frac{T_{m}^2}{2\sigma^2}}\right)
              + \frac{T_{m}}{2} \, \mathrm{erfc}
                \left(-\frac{T_{m}}{\sqrt{2}\sigma}\right)\right] \,,
\end{equation}
which is numerically approximated using week-long sub-intervals. In order to
account for the effects of spatial and seasonal variations of temperature
variability \citep{Seguinot.2013}, $\sigma$ is computed from NARR daily
temperature values from 1979 to 2000 \citep{Mesinger.etal.2006},
including variability associated with the seasonal cycle
itself \citep[Fig.~\ref{fig:atm}; i.e. not as in][]{Seguinot.Rogozhina.2014}.
Degree-day factors for snow and ice melt are derived from
mass-balance measurements on contemporary glaciers from the Coast Mountains and
Rocky Mountains in British Columbia
\citep[Table~\ref{tab:params};][]{Shea.etal.2009}.

@@

Climate forcing driving ice sheet simulations consists of a present-day monthly
climatology,
$\{T_{m0}, P_{m0}\}$, where temperatures are modified by offset time series,
${\Delta}T_{TS}$, and lapse-rate corrections, ${\Delta}T_{LR}$:
\begin{align}
    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{TS}(t)
                    + {\Delta}T_{LR}(t, x, y) \,, \\
    P_m(t, x, y) &= P_{m0}(x, y) \,.
\end{align}

@@

Despite large differences in the input climate forcing
(Fig.~\ref{fig:timeseries}, upper panel), model output presents consistent
features that can be observed across the range of forcing data used. In all
simulations, modelled ice volumes remain relatively low during most of the
glacial cycle, except during two major glacial events which occur between 61.9
and 56.5\,ka during MIS~4, and between 23.2 and 16.8\,ka during MIS~2
(Fig.~\ref{fig:timeseries}, lower panel). An ice volume minimum is
consistently reached between 53.0 and 41.3\,ka during MIS~3. However, the
magnitude and precise timing of these three events depend significantly on the
choice of proxy record used to derive a time-dependent climate forcing
(Table~\ref{tab:extrema}).

Simulations forced by the Greenland ice core palaeo-temperature
records (GRIP, NGRIP) produce the highest variability in modelled ice volume
throughout the last glacial cycle. In contrast, simulations driven by oceanic
(ODP~1012, ODP~1020) and Antarctic (EPICA, Vostok) palaeo-temperature records
generally result in lower ice volume variability throughout the simulation
length, resulting in lower modelled ice volumes during MIS~4 and larger ice
volumes during MIS~3. The NGRIP climate forcing is the only one
that results in a larger ice volume during MIS~4 than during MIS~2.

While simulations driven by the GRIP and the two Antarctic palaeo-temperature
records attain a last ice volume maximum between 19.1 and 16.8\,ka, those
informed by the NGRIP and the two oceanic palaeo-temperature records attain their
maximum ice volumes thousands of years earlier. Moreover, the ODP~1012 run
yields a rapid deglaciation of the modelled area prior to 17\,ka. The ODP~1020
simulation predicts an early maximum in ice volume at 23.2\,ka, followed by
slower deglaciation than modelled using the other palaeo-temperature records.
Finally, model runs forced by the Antarctic palaeo-temperature records result in a
rapid and uninterrupted deglaciation after the LGM, whereas the
simulation driven by the GRIP palaeo-temperature record also results in a rapid
deglaciation but in three steps, separated by two periods of ice sheet regrowth
(Fig.~\ref{fig:timeseries}).

@@

Despite large differences in the timing of attained volume extrema
(Table~\ref{tab:extrema}), all model runs show relatively consistent
patterns of glaciation. During MIS~4, all simulations produce an extensive ice
sheet, covering an area of at least half of that attained during MIS~2
(Table~\ref{tab:extrema}; Fig.~\ref{fig:snapshots}, upper panels).
Corresponding maximum ice volumes also differ significantly between model runs,
and vary between 3.84 and~8.84\,m sea level equivalents (m.~s.l.e.;
Table~\ref{tab:extrema}).

During the MIS~3 ice volume minima, a central ice cap persists over the Skeena
Mountains (Fig.~\ref{fig:snapshots}, middle panels). Although this ice cap is
present in all simulations, its dimensions depend sensitively on the choice of
the applied palaeo-temperature record. Modelled ice volume minima spread over a
wide range between 1.69 and 2.88\,m~s.l.e. (Table~\ref{tab:extrema}).

Modelled ice sheet geometries during the LGM (MIS~2; Fig.~\ref{fig:snapshots},
lower panels) invariably include a ca. 1500\,km-long central divide located
about 3500\,m~a.s.l. along the spine of the Rocky Mountains. Although the
similatity of modelled ice extents is a direct result of the choice of scaling
factors applied to different palaeo-temperature proxy records
(Table~\ref{tab:records}), it is interesting to note that modelled maximum ice
volumes also fall within a tight range of 8.40 to 8.91\,m~s.l.e.
(Table~\ref{tab:extrema}).
